Contracts for unions and intersections are hard for Nickel

### 2021-04-27 paper.tex
Add hyperref
Stress union more than products + stress the importance of custom contracts

\documentclass{article}
\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}

@@

\color{red}Goals of the section

Introduce Nickel, explain the choice of gradual
typing and main design orientations, such as practice-oriented, lightweight,
etc.\vspace{0.5cm}\color{black}

@@

\subsection{Contracts}
\color{red}Goals of the section

A primer on contracts, which in the end are the problematic bit when confronted
with unions.\vspace{0.5cm}

\color{black}

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts.  In this setting, each type constructor of the
static type system - arrows, records, foralls - must have a contract counterpart
that check at run-time that a value is indeed a member of the corresponding
type. Contracts for various extensions of the simply typed $\lambda$-calculus
have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
RootCauseOfBlame, DependentContracts, GradualTypingClasses}, with solutions of
varying complexity.  Amongst all the common extensions to the polymorphic lambda
calculus, one of the most simple and ubiquitous turns out to be one of the most
involved: \emph{union types}.

\subsection{Union types}
\color{red}Goals of the section

Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}\color{black}

Products and unions are fundamental and ubiquitous building blocks of program
data. A product is a compound data that hold several values of potentially
different shape packed together. Dually, a union only holds one value whose
shape is only known to be in a fixed set of alternatives. When nested together,
unions and intersections are able to represent arbitrary tree-like structures,
allowing to model a large class of data.

Unions are also useful for configurations, to represent nullable values in JSON,
plain enumerations, or to support several representations for a field (for
example, a file path could be accepted either as one string of
\lstinline+"/"+-separated strings, or as a list of strings).

@@

\subsection{Tagged versus untagged}
\color{red}Goals of the section

Interpreted languages already need to tag their values, hence bare union are more natural.
It also corresponds better to prior usage in untyped code\vspace{0.5cm}\color{black}

@@

\subsection{Tagged as untagged}
\color{red}Goals of the section

Untagged unions, modulo a slightly smarter typechecker, can encode usual type-safe
tagged unions. They let the user free of the representation though, and encode
more stuff, that's why they are appealing.\vspace{0.5cm}\color{black}

@@

\subsection{Union and intersections}
\color{red}Goals of the section

Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}\color{black}

>>>>

\documentclass{article}
\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{hyperref}

@@

{\color{red}Goals of the section

Introduce Nickel, explain the choice of gradual
typing and main design orientations, such as practice-oriented, lightweight,
etc.\vspace{0.5cm}}

@@

\subsection{Contracts}
{\color{red}Goals of the section

A primer on contracts, which in the end are the problematic bit when confronted
with unions.\vspace{0.5cm}}

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts. In "first-class contracts", we include the
ability to write user-defined contracts. Because built-in types are too
simplistic for the purpose of data validation, the ability of writing a small
validation function for, say, an URL, and having it immediately available as a
contract that can be composed with other type constructors, is in our
opinion one of the main appeal of Nickel contracts system. In this hybrid
system, each type constructor of the static type system - arrows, records,
foralls - must have a contract counterpart that checks at run-time that a value
is indeed a member of the corresponding type. Contracts for various extensions
of the simply typed $\lambda$-calculus have been investigated \cite{BlameForAll,
    KeilThiemannUnionIntersection, RootCauseOfBlame, DependentContracts,
GradualTypingClasses}, with solutions of varying complexity.  Amongst all the
common extensions to the polymorphic lambda calculus, one of the most useful and
widespread turns out to be one of the most challenging: \emph{union types}.

\subsection{Union types}
{\color{red}Goals of the section

Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}}

Unions are fundamental and ubiquitous building blocks of program data. A union
only holds one value whose shape is only known to be in a fixed set of
alternatives: for eample, either a boolean or a string. Wuen used in combination
with products, they can represent arbitrary tree-like structures, allowing to
model a large class of data.

Unions are also useful for configurations:
\begin{itemize}
    \item To represent nullable values in JSON
    \item For plain enumerations. For, the enabled protocols of a data server
        would be either \lstinline+HTTP+, \lstinline+FTP+ or \lstinline+SFTP+.
    \item To support different representations for a field. A file path could be
        accepted either as one string of \lstinline+"/"+-separated strings, or
        as a list of strings.
\end{itemize}

@@

\subsection{Tagged versus untagged}
{\color{red}Goals of the section

Interpreted languages already need to tag their values, hence bare union are more natural.
It also corresponds better to prior usage in untyped code\vspace{0.5cm}}

@@

\subsection{Tagged as untagged}
{\color{red}Goals of the section

Untagged unions, modulo a slightly smarter typechecker, can encode usual type-safe
tagged unions. They let the user free of the representation though, and encode
more stuff, that's why they are appealing.\vspace{0.5cm}}

@@

\subsection{Union and intersections}
{\color{red}Goals of the section

Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}}
